Visualizing Multi-way Contingency Tables

### 2005-11-16 trunk/vcd/paper/hsc.Rnw
-bug fixes in pairs plot -sieve plots now use HCL colors -`labeling' argument now also accepts a logical
-hsc-paper revised



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@498 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title*{Visualizing Multi-Way Contingency Tables}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{David Meyer\inst{1}\and
Achim Zeileis\inst{2}\and
Kurt Hornik\inst{2}}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Department of Information Systems and Process Management
\and Department of Statistics and Mathematics\\Wirtschaftsuniversität
Wien\\Augasse 2--6, 1090 Vienna, Austria
\texttt{\{David.Meyer,Achim.Zeileis,Kurt.Hornik\}@wu-wien.ac.at}}
\maketitle

@@

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
statistical methods such as independence tests and
log-linear models. The analysis quickly 
becomes complex with increasing dimensionality.
One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. All three have been
extended over the last two decades, and implementations exist in many
statistical environments.

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plot} are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

Over the years, extensions to these techniques were mainly focused on
five aspects:

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \cite{vcd:hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).

@@

Our contribution will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section \ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Sect.~\ref{sec:concl} concludes the chapter.

@@

Throughout these section, our examples will be based on
the \data{hospital} data \cite{vcd:haberman:1974} (see Tab.~\ref{tab:hospital}).

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||rrr|r|}
\hline
 & Length of stay &&&\\
 & 2--9 days & 10--19 days & more than 20 days & TOTAL \\
\hline\hline
Visit frequency &&&&\\
Regular & 43 & 16 & 3 & 62 \\
Less than monthly & 6 & 11 & 10 & 27 \\
Never & 9 & 18 & 16 & 43 \\\hline
TOTAL & 58 & 45 & 29 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data.}
\label{tab:hospital}
\end{center}
\end{table}

\noindent The table relates the length of stay (in days) in hospital and
the visit freqency for mental patients, 
giving evidence to the fact that the longer the length of 
stay in hospital, the less frequent the visits.

Although far from optimal, contingency tables are frequently visualized using
grouped bar charts (see Fig.~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<grouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
barplot(hospital, legend = rownames(hospital), beside = TRUE, 
        xlab = "Length of stay", ylab = "Number of patients")
@
\caption{Barplot for the \data{hospital} data.}
\label{fig:grouped}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Regular","","Less than monthly","","Never"), 
              xlab = "Length of stay", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-barplot for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square. Since 
chapter XXX in this book \emph{\bf (Reference to the contribution of Heike Hofmann)} 
is entirely devoted to mosaic plots, in the following we only describe
the main idea. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose `Length of stay' with vertical splits 
(see Fig.~\ref{fig:mosaic1fig}).

@@

\noindent The
result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which some authors \cite{vcd:hummel:1996}
call a \emph{spine plot}. From this plot, we see that the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still depict the marginal distribution of `Length of stay', %'
and additionally, the visit frequency \emph{given} each category of `Length of stay'. %'
As we will see in the next section, the grid would be regular
if the two variables were independent. Clearly, compared
to a length of stay of 10--19 days, more patients get regular visits for stays from 2--9 
days, and conversely, less patients get regular visits for stays for more than 20 days.
For patients that get no visits, the pattern is inversed.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 days.
It seems apparent that the distribution of `Length of stay' %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually).

@@

When we try to explain data, we suppose the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common hypotheses are

\begin{enumerate}
\item independence of the two variables, and
\item homogeneity of one variable among the strata defined by
the second.
\end{enumerate}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||rrr|r|}
\hline
 & Length of stay &&&\\
 & 2--9 & 10--19 & 20+ & TOTAL \\
\hline\hline
Visit frequency &&&&\\
Regular & 27.24 & 21.14 & 13.62 & 62 \\
Less than monthly & 11.86 & 9.20 & 5.93 & 27 \\
Never & 18.89 & 14.66 & 9.45 & 43 \\\hline
TOTAL & 58.00 & 45.00 & 29.00 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data---expected values.}
\label{tab:hospital2}
\end{center}
\end{table}

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. In
Fig.~\ref{fig:sieve1fig}, we can see such a plot in which,
additionally, a number of squares equal to the corresponding cell entry has
been added to each tile. We see that the grid formed by the tiles is completely
regular. The added rectangles in the tiles become more useful when they 
are not based on the expected but the observed values
instead (see Fig.~\ref{fig:sieve2fig}). Such a \emph{sieve plot} implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the plot reveals indeed a negative association of 
the two variables since the density of the rectangles is marked along the secondary
diagonal, giving evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<sieve1fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, sievetype = "expected", pop = FALSE)
labeling_cells(text = round(independence_table(t(hospital)),2), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Expected values for the \data{hospital} data.}
\label{fig:sieve1fig}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<sieve2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "dark gray"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sieve2fig}
\end{center}
\end{figure}

@@

\noindent that are standardized raw residuals. They are best visualized by
association plots \cite{vcd:Cohen:1980}:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. In available
implementations, the sign of a residual is often redundantly
coded, e.g., by the rectangle's color (see Sect.~\ref{sec:shadings}) %'
and its position relative to the baseline. Figure \ref{fig:assoc}
shows the association plot for the hospital data. 
Consistent with the corresponding mosaic and sieve plots, 
we clearly see that too many (too less) patients that stay from 
2 to 9 (more than 20) days get visited regularly
than would be expected under the null of independence, and that this pattern is 
reversed for patients never visited.

@@

\section{Residual-based Shadings}
\label{sec:extensions}

As introduced in the previous section for association plots, the
investigation of residuals from a posited independence model is of
major interest in analyzing contingency tables. In the following, we
will demonstrate how the use of colors can greatly facilitate the
detection of interesting patterns.

>>>>

\title*{Visualizing Multi-way Contingency Tables}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{David Meyer\inst{1}\and
Achim Zeileis\inst{2}\and
Kurt Hornik\inst{2}}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Department of Information Systems and Process Management
\and Department of Statistics and Mathematics\\Wirtschaftsuniversität
Wien\\Augasse 2--6, 1090 Vienna, Austria
\texttt{\{David.Meyer,Achim.Zeileis,Kurt.Hornik\}@wu-wien.ac.at}}
\maketitle

@@

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. They have been
extended over the last two decades, and implementations exist in many
statistical environments.

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

Over the years, extensions to these techniques have mainly been focused on
the following aspects:

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}).

@@

This chapter will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Sect.~\ref{sec:concl} concludes the chapter.

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} (see Tab.~\ref{tab:hospital}).

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
\hline
Regular & 43 & 16 & 3 & 62 \\
Less than monthly & 6 & 11 & 10 & 27 \\
Never & 9 & 18 & 16 & 43 \\\hline
$\Sigma$ & 58 & 45 & 29 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data.}
\label{tab:hospital}
\end{center}
\end{table}

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the freqency of visits. The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Tab.~\ref{tab:colhospital}).

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|}
\hline
 & \multicolumn{3}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ \\
\hline
Regular & 0.74 & 0.36 & 0.10  \\
Less than monthly & 0.10 & 0.24 & 0.35  \\
Never & 0.16 & 0.40 & 0.55  \\\hline
$\Sigma$ & 1.00 & 1.00 & 1.00 \\
\hline
\end{tabular}
\caption{The \data{hospital} data, corrected for the column margin.}
\label{tab:colhospital}
\end{center}
\end{table}

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab~.\ref{tab:rowhospital}), 
it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
\hline
Regular & 0.69 & 0.26 & 0.05 & 1.00 \\
Less than monthly & 0.22 & 0.41 & 0.37 & 1.00 \\
Never & 0.21 & 0.42 & 0.37 & 1.00 \\
\hline
\end{tabular}
\caption{The \data{hospital} data, corrected for the row margin.}
\label{tab:rowhospital}
\end{center}
\end{table}

Although far from optimal, contingency tables are frequently visualized using
grouped bar plots (see Fig.~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<grouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
barplot(hospital, legend = rownames(hospital), beside = TRUE, 
        xlab = "Length of stay (years)", ylab = "Number of patients")
@
\caption{Bar plot for the \data{hospital} data.}
\label{fig:grouped}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Regular","","Less than monthly","","Never"), 
              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-bar chart for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is inversed.

@@

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable in (hypothesized) causal relationships.

@@

When we try to explain data, we assume the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common hypotheses \cite{vcd:Agresti:2002} are

\begin{enumerate}
\item independence of the two variables
\item homogeneity of one variable among the strata defined by
the second.
\end{enumerate}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} & ~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
\hline
Regular & 27.24 & 21.14 & 13.62 & 62 \\
Less than monthly & 11.86 & 9.20 & 5.93 & 27 \\
Never & 18.89 & 14.66 & 9.45 & 43 \\\hline
$\Sigma$ & 58.00 & 45.00 & 29.00 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data---expected values.}
\label{tab:hospital2}
\end{center}
\end{table}

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Fig.~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "dark gray"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sievefig}
\end{center}
\end{figure}

@@

\noindent which are standardized raw residuals. In an 
association plot \cite{vcd:Cohen:1980}, each cell is represented 
by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The sign is visualized by
its position relative to the baseline (upward tiles for positive, and downward 
tiles for negative residuals). Figure~\ref{fig:assoc}
shows the association plot for the hospital data. 
Consistent with the corresponding mosaic and sieve plots, 
we clearly see that too many (too few) patients that stay from 
2 to 9 (more than 19) years get visited regularly
than would be expected under the null of independence, and that this pattern is 
reversed for patients visited less than monthly and that get never visited.

@@

\subsection{Summary}

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If causal
relationships between the variables are assumed, the determining variables
should be used first for splitting since they will dominate the
plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They are best used to detect dependency
patterns, especially for ordinal variables. Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary concern.

\section{Using colors for Residual-based Shadings}
\label{sec:extensions}

As introduced in the previous section for association plots, the
investigation of residuals from a posited independence model is of
major interest in analyzing contingency tables. In the following, we
will demonstrate how the use of colors can greatly facilitate the
detection of interesting patterns. Before, we start with some general 
remarks on colors and color palettes.

\subsection{A Note on Colors and Color Palettes}
\label{sec:colors}

The plots introduced in the previous section are basically composed of
tiles whose areas represent characteristics derived from the
contingency tables---observed and expected frequencies in the case of
mosaic and sieve plots, or residuals as visualized by association
plots. When using color for these tiles, it is imperative
to choose the right color palettes, derived from suitable color
spaces. Apart from esthetic considerations, wrongly chosen colors
might serously affect the analysis: For example,

\begin{itemize}
\item Using high-chroma colors for large
  areas tend to produce after-image effects which can be
  distracting \cite{vcd:Ihaka:2003}.
\item Lighter colors tend to make areas look larger than darker
  colors, therefore using colors with unequal luminance values  make it
  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
\item Color palettes derived from
  non-uniform color spaces may contain unbalanced colors with respect to
  their colorfulness or lightness. When tiles are shaded using such a
  palette, some of them might appear more important than others in an
  uncontrolled way.
\end{itemize}

% gamma correction for displays.

\noindent Implementations are often based on so-called
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions: the hue (dominant wavelength in the spectrum), 
the saturation (`colorfulness' or `pureness'), and 
the value (`brightness', amount of gray). These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red over a neutral white
to a saturated blue. Although the palette should be balanced with
respect to colorfulness and lightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@ 
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for 
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.}
\label{fig:diverging}
\end{center}
\end{figure}
